The Evolution of Phenotypically Invariant Genetic Networks

### 2017-12-07 kalman_walk_paper.tex
edits

Nonidentifiability is a major barrier to mechanistic understanding of real systems, 
but viewed from another angle,
this concept can provide a starting point for thinking about externally equivalent systems
-- systems that evolution can explore, so long as the parameters and structures can be realized biologically.
These functional symmetries manifest in convergent and parallel evolution, 
as well as \emph{developmental system drift}: the observation that
macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
\citep{kimura1981possibility, true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, lavoie2010evolutionary, vierstra2014mouse, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription}.
It has also been shown theoretically that substantial genetic variation for a polygenic trait can be maintained
by migration between geographical regions having different genetic bases for the same, optimal trait value \citep{phillips1996maintenance}.

@@

It is not a new observations that there is often more than one way to do the same thing,
and that this may lead to speciation.
For instance, \citet{tulchinsky2014hybrid} (and similarly \citet{khatri2015simple}) looked for hybrid incompatibility
following regulatory sequence drift,
and \citet{porter2002speciation} studied speciation in populations expressing a simple regulatory cascade.
Both found rapid speciation following directional selection,
but only equivocal support for speciation under models of purely neutral drift.
Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
Our model differs from previous work in that we use linear systems theory
to both explore a much richer class of regulatory network models,
and provide analytical expectations in large populations with complex phenotypes
that would be inaccessible to population simulations.
%REFERENCE: Barton's paper (PETER). \citep{barton2010natural}

@@

\section*{Acknowledgements}
We would like to thank Sergey Nuzhdin, Stevan Arnold, Erik Lundgren, and Hossein Asgharian for valuable discussion.

\plr{Josh: please fix capitalization in the references.  To do this, anything that needs to be capitalized gets put in curly braces, which tells bibtex to render the capitalization 'as-is'.}

>>>>

Nonidentifiability is a major barrier to mechanistic understanding of real systems, 
but viewed from another angle,
this concept can provide a starting point for thinking about externally equivalent systems
-- systems that evolution can explore, so long as the parameters and structures can be realized biologically.
These functional symmetries manifest in convergent and parallel evolution, 
as well as \emph{developmental system drift}: the observation that
macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
\citep{kimura1981possibility, true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, lavoie2010evolutionary, vierstra2014mouse, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription}.
%It has also been shown theoretically that substantial genetic variation for a polygenic trait can be maintained
%by migration between geographical regions having different genetic bases for the same, optimal trait value \citep{phillips1996maintenance}.

@@

It is not a new observations that there is often more than one way to do the same thing,
and that this may lead to speciation.
For instance, \citet{tulchinsky2014hybrid}
%(and similarly \citet{khatri2015simple})
looked for hybrid incompatibility
following regulatory sequence drift,
and \citet{porter2002speciation} studied speciation in populations expressing a simple regulatory cascade.
Both found rapid speciation following directional selection,
but only equivocal support for speciation under models of purely neutral drift.
%Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
Our model differs from previous work in that we use linear systems theory
to both explore a much richer class of regulatory network models,
and provide analytical expectations in large populations with complex phenotypes
that would be inaccessible to population simulations.
%REFERENCE: Barton's paper (PETER). \citep{barton2010natural}

@@

\section*{Acknowledgements}
We would like to thank Sergey Nuzhdin, Stevan Arnold, Erik Lundgren, and Hossein Asgharian for valuable discussion.
%
%\plr{Josh: please fix capitalization in the references.  To do this, anything that needs to be capitalized gets put in curly braces, which tells bibtex to render the capitalization 'as-is'.}
